---
title: "Analysis Walkthrough"
format:
  html:
    css: styles.css
    embed-resources: true
    code-fold: false
    page-layout: full
    fig_caption: true
    toc: false
---

::: {#page-banner}
<h1 class="page-title">

Analysis Walkthrough

</h1>
:::

:::: narrow-page
## About the Data

This dataset includes 289 deaths in [West Virginia Division of Corrections and Rehabilitation (DCR)](https://dcr.wv.gov/Pages/default.aspx) custody from January 2020 through March 2025. It was obtained by the [West Virginia Center on Budget and Policy](https://wvpolicy.org/) through a FOIA request filed by senior policy analyst [Sara Whitaker](https://wvpolicy.org/team/).

Each row represents a death, with key fields including:

-   `date_of_death` and `booking_date`
-   `age`, `race`, `gender`
-   `facility` and `offender_status` (e.g., pretrial vs. sentenced)
-   `manner_of_death` and `cause_of_death`

DCR’s annual reports do not publish this level of detail. Jail deaths are excluded, and no information is provided by facility, demographics, or cause. This dataset allows for deeper analysis of where, when, and how people are dying in custody.

## Exploratory Data Analysis

In this section, we’ll begin exploring:

-   How many deaths occurred each year?
-   Which facilities have the highest number of deaths?
-   How do deaths vary by custody status (pretrial vs. sentenced)?
-   Are certain groups disproportionately affected?
-   Are facilities with high mortality rates experiencing overcrowding?

We’ll also examine patterns in cause of death and compare counts to what DCR reports in its official publications.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  include = TRUE
)

source("R/setup.R")
source("R/utils_helpers.R")

# Load WV Center on Budget and Policy data
wvcbp_deaths <- readRDS("data/wvcbp_deaths_in_custody.rds")

df <- prepare_deaths_data(wvcbp_deaths)
df_wo_2025 <- df |> filter(year < 2025)
```

## Annual Trends

This section tracks the number of reported deaths by calendar year.

```{r, asis=TRUE}
# Deaths per year for display
annual_counts <- df |>
  count(year) |>
  arrange(year)

# Get values for 2024 and Jan–Mar 2025
deaths_2024 <- annual_counts |> filter(year == 2024) |> pull(n)
deaths_2025_ytd <- df |> filter(year == 2025, lubridate::month(date_of_death) <= 3) |> nrow()
```

While deaths have been gradually declining in recent years, there were still `r deaths_2024` deaths in 2024. In the first three months of 2025 alone, there have already been `r deaths_2025_ytd` deaths reported. This raises questions about whether recent policy changes or population shifts are reducing risk—or if conditions remain just as deadly.

```{r}
plot_deaths_by_year(df_wo_2025)
```

## Demographics

Understanding who is dying in custody is critical. Demographic breakdowns help show whether certain groups—by race, gender, or age—are disproportionately represented in deaths.

A disparities analysis is in progress. For example, assessing whether certain racial groups experience higher death rates relative to their representation in the incarcerated population. This includes calculating relative rate indices using population baselines from facility-level and systemwide data where available.

::: {.panel-tabset .nav-pills}

### Race

`r generate_summary_sentence(df_wo_2025, "race")`

```{r}
plot_deaths_by_var(df_wo_2025, "race") |>
  hc_chart(height = 350)
```

### Gender

`r generate_summary_sentence(df_wo_2025, "gender")`

```{r}
plot_deaths_by_var(df_wo_2025, "gender") |>
  hc_chart(height = 350)
```

### Age

```{r}
# Filter valid age data
age_summary <- df_wo_2025 |>
  filter(!is.na(age_at_time_of_death))

# Counts
over_55 <- age_summary |> filter(age_at_time_of_death >= 55) |> nrow()
under_55 <- age_summary |> filter(age_at_time_of_death < 55) |> nrow()
total_age_count <- nrow(age_summary)

# Percentages
pct_over_55 <- round(100 * over_55 / total_age_count, 1)
pct_under_55 <- round(100 * under_55 / total_age_count, 1)

# Average age
avg_age <- round(mean(age_summary$age_at_time_of_death), 1)
```

`r generate_summary_sentence(df_wo_2025, "age")`

`r under_55` people who died in custody from 2020 to 2024 were under the age of 55 — that’s `r pct_under_55`% of all deaths. `r over_55` people (or `r pct_over_55`%) were 55 or older.

The average age at time of death was `r avg_age` years old. For comparison, the average life expectancy in the U.S. is about 76 years.

```{r}
plot_deaths_by_var(df_wo_2025, "age") |>
  hc_chart(height = 350)
```
:::

## Offender Status (Jail)

Many people who die in jail custody are being held pretrial, meaning they have not been convicted of a crime. Despite this, they are housed in facilities operating far [beyond their intended capacity](https://wvpolicy.org/officials-deny-overcrowding-but-29-of-people-in-jail-assigned-to-sleep-on-the-floor/). In some jails—like North Central Regional Jail—population levels have exceeded twice the design limit.

As Sara Whitaker has documented, people in these overcapacity facilities are often forced to sleep on mats on the floor, with limited access to basic hygiene and medical care. These conditions are not temporary. In some jails, overcapacity has been the norm for years, raising serious concerns about safety, dignity, and the state’s duty of care.

```{r}
df_jail <- df_wo_2025 |> filter(type == "Jail")
```

`r generate_summary_sentence(df_jail, "offender_status")`

```{r}
plot_deaths_by_var(df_jail, "offender_status")
```

## Offender Status (Prison)

This chart examines deaths in prison settings, where most individuals are serving post-conviction sentences. Comparing jail and prison trends can highlight differences in medical access, length of stay, and population health.

```{r, asis=TRUE}
df_prison <- df_wo_2025 |> filter(type == "Prison")
```

`r generate_summary_sentence(df_prison, "offender_status")`

```{r}
plot_deaths_by_var(df_prison, "offender_status") 
```

## Cause of Death

`r generate_summary_sentence(df_wo_2025, "cause_of_death")`
`r generate_additional_cod_note(df_wo_2025)` These vague labels make it difficult to assess medical conditions or facility care quality.

```{r}
plot_deaths_by_var(df_wo_2025, "cause_of_death")
```

## Manner of Death 

`r generate_summary_sentence(df_wo_2025, "manner_of_death")` Classifying a death as "natural" doesn’t necessarily mean it was unavoidable. Further review is needed to assess access to medical care and facility conditions.

```{r}
plot_deaths_by_var(df_wo_2025, "manner_of_death")
```

## Facilities With the Most Deaths

Some facilities account for a disproportionately high number of deaths.

Currently investigating whether these patterns may be related to conditions such as overcrowding or staffing shortages. As part of the next phase of this analysis, I will calculate mortality rates, using population data.

```{r}
top_facilities <- df_wo_2025 |>
  group_by(type, facility_name) |>
  summarise(total_deaths = n(), .groups = "drop") |>
  arrange(desc(total_deaths))

# Get top 2 jails and top 2 prisons
top_prisons <- top_facilities |> filter(type == "Prison") |> slice_max(total_deaths, n = 2)
top_jails <- top_facilities |> filter(type == "Jail") |> slice_max(total_deaths, n = 2)

# Format sentence
sentence <- paste0(
  "From 2020 to 2024, the prisons with the most deaths were ",
  top_prisons$facility_name[1], " and ", top_prisons$facility_name[2],
  ". The jails with the most deaths were ",
  top_jails$facility_name[1], " and ", top_jails$facility_name[2], "."
)
```
`r sentence` A forthcoming analysis will examine whether these facilities were operating over capacity and how that may relate to deaths.

```{r}
reactable(
  top_facilities,
  theme = reactable_theme(),
  filterable = FALSE,
  pagination = FALSE,
  defaultColDef = colDef(na = "–"),
  columns = list(
    type = colDef(name = "Facility Type"),
    facility_name = colDef(name = "Facility Name"),
    total_deaths = colDef(name = "Total Deaths")
  )
)
```

## Net Change in Deaths: 2020 vs 2024

```{r}
calculate_net_change <- function(df_wo_2025, start_year = 2020, end_year = 2024) {
  df_wo_2025 |>
    filter(year %in% c(start_year, end_year)) |>
    count(facility_name, year) |>
    pivot_wider(names_from = year, values_from = n, values_fill = 0) |>
    mutate(net_change = !!sym(as.character(end_year)) - !!sym(as.character(start_year))) |>
    arrange(desc(net_change))
}

net_change_df <- calculate_net_change(df) |> 
  arrange(desc(net_change))
reactable(
  net_change_df,
  theme = reactable_theme(),
  filterable = FALSE,
  pagination = FALSE,
  defaultColDef = colDef(na = "–"),
  columns = list(
    facility_name = colDef(name = "Facility Name"),
    net_change = colDef(name = "Net Change")
  )
)
```

## Next Steps

As part of the next phase of this analysis, I plan to calculate mortality rates by facility type and location. This will allow for comparisons that account for population size. For example, a facility with 5 deaths and an average daily population of 100 has a far higher death rate than one with 10 deaths and 1,000 people. These rates will help identify facilities where the risk of dying is especially high, regardless of total population.

Projects like the [UCLA Behind Bars Data Project](https://uclaprisondata.org/) use similar methods to track death rates across systems—often using reported custody populations as denominators.

::::
