---
title: "Download Data"
format:
  html:
    body-class: wide-page
    css: styles.css
    embed-resources: true
    code-fold: false
    page-layout: full
    fig_caption: true
    toc: false
---

::: {#page-banner}
<h1 class="page-title">
Download Data
</h1>
:::
<div class="narrow-page">
<br>
<p>This dataset is based on records obtained through FOIA requests to the <strong>West Virginia Division of Corrections and Rehabilitation</strong> and compiled by the <strong>West Virginia Center on Budget and Policy</strong> (WVCBP). It was published in March 2025 as part of their reporting on deaths in state custody.</p>
</div>

::: {#wide-download-page}
<a href="data/wvcbp_deaths_in_custody.csv" class="button">Download CSV</a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  include = TRUE
)
```

```{r}
source("R/setup.R")
source("R/utils_helpers.R")

# Load WV Center on Budget and Policy
wvcbp_deaths <- readRDS("data/wvcbp_deaths_in_custody.rds")

full_names <- c(
  date_of_death = "Date of Death",
  booking_date = "Booking Date",
  race = "Race",
  gender = "Gender",
  age_at_time_of_death = "Age",
  facility = "Facility Code",
  manner_of_death = "Manner of Death",
  cause_of_death = "Cause of Death",
  offender_status = "Offender Status",
  county = "County",
  source_table = "Source Table",
  dob = "DOB",
  case_number = "Case #",
  facility_name = "Facility",
  type = "Facility Type"
)

# Custom widths for specific columns
custom_col_defs <- list(
  date_of_death = colDef(name = "Date of Death", width = 100),
  booking_date = colDef(name = "Booking Date", width = 100),
  age_at_time_of_death = colDef(name = "Age", width = 60),
  facility_name = colDef(name = "Facility", width = 220),
  cause_of_death = colDef(name = "Cause of Death", width = 180),
  
  offender_status = colDef(name = "Offender Status", width = 190),
  county = colDef(name = "County", width = 180),
  dob = colDef(name = "DOB", width = 100),
  case_number = colDef(name = "Case #", width = 100)
)

# Build full column definitions: use custom if available, else default
column_defs <- lapply(names(full_names), function(col) {
  # Start with default colDef
  base <- colDef()
  
  # If there's a custom colDef, merge its settings into base
  if (col %in% names(custom_col_defs)) {
    base <- modifyList(base, custom_col_defs[[col]])
  }

  # Always apply the human-readable name
  base$name <- full_names[[col]]
  base
}) |> setNames(names(full_names))


# Final reactable call
reactable(
  wvcbp_deaths |> select(-c(source_table, facility)) |> 
    select(date_of_death, booking_date, race, gender, age_at_time_of_death, facility_name, type, everything()),
  theme = reactable_theme(),
  columns = column_defs,
  searchable = FALSE,
  filterable = TRUE,
  pagination = TRUE,
  defaultColDef = colDef(
    na = "–",
    width = 80
  ),
  bordered = FALSE,
  highlight = TRUE,
  striped = FALSE
)
```
:::

<br>
**Data Notes:**

<p>You can view the original source file here:  
<a href="https://wvpolicy.org/wp-content/uploads/2025/03/DCR-Deaths-in-Custody-Jan-2020-to-March-2025.pdf" target="_blank">DCR Deaths in Custody: Jan 2020–March 2025 (PDF)</a>.</p>

<p>The dataset was released publicly in this article:  
<a href="https://wvpolicy.org/are-fewer-people-dying-in-west-virginia-jails-or-is-the-state-simply-not-counting/" target="_blank">Are fewer people dying in West Virginia jails, or is the state simply not counting?</a></p>

<p>Here are some FOIA request documents cited or referenced in that reporting:</p>

<ul>
  <li>
    <a href="https://wvpolicy.org/wp-content/uploads/2024/10/2024-09-25-DCR-Response-to-Whitaker-Sara-N.-FOIA-request-inmate-count-09-03-24-deaths-PDs-25.pdf" target="_blank">
      September 2024 FOIA response – death log from August 2024
    </a>
  </li>
  <li>
    <a href="https://wvpolicy.org/wp-content/uploads/2024/04/2024.01.16-DCR-Response-to-Tony-Mike-Gazette-FOIA-request-DCR-inmate-deaths-2009-present-1.pdf" target="_blank">
      January 2024 FOIA response – deaths in custody (2009–present)
    </a>
  </li>
  <li>
    <a href="https://wvpolicy.org/wp-content/uploads/2025/02/2025.02.14-DHS-FOIA-Response-re-Inspector-General-reports.pdf" target="_blank">
      February 2025 FOIA response – Inspector General reports
    </a>
  </li>
</ul>